---
title: "Assessment"
author: "Huw Riley"
output:
  html_document:
    theme: darkly
    toc: true
    number_sections: true
    code_folding: hide
    fig_caption: true
---

```{r setup, include=FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)
```

## Introduction

  Paracetamol is a very common drug used in Scotland for pain relief. While it can be obtained over the counter from supermarkets and pharmacies, it is still often prescribed in hospitals and the community. When obtained without prescription, paracetamol costs the patient on average £2.43 per 100 tablets while prescribing paracetamol to patients costs the NHS £14.18 for the same quantity[1]. This is due to the costs from dispensing the prescription from the pharmacy as well as the costs of professionals to write the prescriptions[2]  Since the NHS spends around £136 million per year on prescribing medications that can be bought over the counter for cheaper[1], it is important we look to see if there are demographics where these are being prescribed more to identify how we can save the NHS money that could be better spent in other ways. Once these trends are identified, this could be used to help inform changes to prescribing guidelines, improving future clinical practise. 
  Through looking at prescribing data from Scotland in 2024, I will be looking to see if there are trends in how the paracetamol prescriptions are divided over the population as well as how they change across the year. Firstly, I will look at which health boards prescribe the most paracetamol before looking at how the prescription of paracetamol changes across the year. Following this, I will look to see if there is a correlation between amount of elderly people within a population and paracetamol prescriptions. This is important as I would expect areas with more elderly people to prescribe more paracetamol due to elderly people being more likely to have comorbidities. Finally, I will use the Scottish Index of Multiple Deprivation (SIMD) to look at whether there is a correlation between how deprived an area is and paracetamol prescriptions. I would expect there to be greater number of prescriptions in more deprived areas, possible due to lack of means to purchase paracetamol themselves, whether through being unable to afford it or unable to get to a place to purchase it.

## Methods

  For my analysis, I will be looking at 500mg doses of paracetamol only as these are the most commonly sold measure of paracetamol in supermarkets and therefore the most fair comparison between the two. I will be using prescription data from 2024 as it is the most recent complete year available to me. I am only looking at data of prescriptions in the community, leaving out hospital prescriptions, as it is unrealistic to expect hospitalized patients to purchase their own paracetamol. Therefore they must rely on prescriptions for paracetamol so these prescriptions would be harder to reduce.


```{r library, error = FALSE, warnimg = FALSE}
library(tidyverse); theme_set(theme_linedraw()) 
library(janitor) #Gives Clean_names function
library(purrr) #allows me to use mapping for loading data
library(sf) #allows me to create maps
library(gt) #allows me to create better tables
library(ggiraph) #allows me to create an interactive map
```


```{r Function Creation}
url_loading <- function(url) {read_csv(url, show_col_types = FALSE) %>% 
    clean_names()} #creates a function to load all csv files in a uniform way

data_cleaning <- function(date, url) { #function to clean all 12 prescription files
    url_loading(url)%>% 
    mutate(dmd_code = as.character(dmd_code)) %>% #standardises dmd_code so always character
    filter(str_starts(bnf_item_description, "PARACETAMOL 500MG")) %>% #only looking at 500mg so removes anything else
    mutate(date = as.Date(paste0(`date`, "/01"), format = "%m/%Y/%d"))} #puts date in correct format
```
```{r Prescription Information}
dates <- sprintf("%02d/2024", 1:12) #creates a dataframe with dates to add to the loaded data
urls <- read_lines("links_clean.csv") #reads the csv with urls of all prescription data (stored in docs)
prescriptions_2024 <-map2_dfr(dates, urls, data_cleaning) 
```
```{r Healthboard Information}
hb_names <- url_loading("https://www.opendata.nhs.scot/dataset/9f942fdb-e59e-44f5-b534-d6e17229cc7b/resource/652ff726-e676-4a20-abda-435b98dd7bdc/download/hb14_hb19.csv") %>% #adding health board names
  select(hb, hb_name) #selects only the useful columns
prescriptions_with_boards <- left_join(prescriptions_2024, hb_names, by =c("hbt"="hb"))

hb_population <- url_loading("https://www.opendata.nhs.scot/dataset/7f010430-6ce1-4813-b25c-f7f335bdc4dc/resource/27a72cc8-d6d8-430c-8b4f-3109a9ceadb1/download/hb2019_pop_est_03102025.csv") %>%  #adding health board populations
  filter(year == "2024", sex == "All") #gives us only 2024 populations and removes unneeded gender split
full_prescriptions_info <- left_join(prescriptions_with_boards, hb_population, by =c("hbt"="hb"))
```


```{r}
per_100k <- function(group){ #prepares data with totals and per 100,000 per supplied group
  full_prescriptions_info %>% 
  filter(!is.na(hb_name)) %>% 
  group_by(across(all_of(group))) %>% 
  summarise(number_of_paid_items = sum(number_of_paid_items), population = mean(all_ages)) %>% 
  mutate(items_per_100k = number_of_paid_items / population * 100000)} 

Prescriptions_per_board <- per_100k("hb_name") %>% #creating table of yearly prescriptions per health board
  arrange(desc(items_per_100k)) %>% 
   rename(Healthboard = hb_name, Prescriptions = number_of_paid_items, "Figure 1: Prescriptions Per 100,000 People" = items_per_100k) %>% 
  gt() %>% 
  tab_header(title = "Table of Yearly Prescriptions of Paracetamol Per year", subtitle = "Sorted by Prescriptions Per 100,000 People in Descending Order") %>% 
  fmt_number(decimals = 0, use_seps = TRUE) #removes decimals and adds commas to numbers
Prescriptions_per_board#displaying table
```
## Results

### Prescriptions Per Healthboard

  From the results of figure 1, we can see that the health board that prescribes the most paracetamol relative to population is Western Isles. They prescribed 74,581 per 100,000 people which is significantly more than Dumfries and Galloway which was second with 52,307. Beyond the top health board, the next 12 are all fairly evenly spread between 35,600 and 54,000 with the final board, Lothian, being significantly the lowest with 27,934. It is worth noting that 3 out of the top 4 health boards are on islands separate to the Scottish mainland so it is possible that this is a potential cause of higher paracetamol prescriptions. This is perhaps due to people in the island boards being more rural and therefore having less access to supermarkets to purchase their own paracetamol and therefore they require the gp to prescribe it for them. This could be further supported by the fact that health boards containing Scotlands two largest cities, Edinburgh and Glasgow, having the lowest and fourth lowest prescriptions per 100,000 people. This supports the hypothesis as these large cities are likely to have more people living within close proximity of multiple supermarkets.
  However, it may be worth looking at which health boards have the highest overall paracetamol prescriptions regardless of population as these are the areas where reducing the amount of prescriptions could have the greatest overall financial impact. In this case, Greater Glasgow and Clyde has by far the most paracetamol prescriptions, with 461,769 per year, followed by Lothian, Lanarkshire and Grampian, all of which have over 200,000 per year. This in contrast to the island boards, with each having less than 20,000 prescriptions per year. This means that even if you cut all prescriptions from one of these health boards, this would only be equivalent to cutting at most 4.2% of Greater Glasgow and Clyde's prescriptions ((19,406/461,769)*100).

```{r monthly graph}
paracetamol_per_board <- per_100k(c("hb_name", "date")) %>% #creating graph of monthly prescribtions per board
  ggplot(aes(x = date, y = items_per_100k)) +
  geom_line() +
  facet_wrap(~hb_name) +
  ggtitle("Figure 2: Total Paracetamol prescriptions per 100,000 population by health board in Scotland (2024)") +
  xlab("Date") +
  ylab("Number of Prescriptions per 100,000") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1), plot.title = element_text(size=11.5), strip.text= element_text(size = 7)) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, NA)) #allows comparison between health boards
paracetamol_per_board #displays graph
```

### How prescriptions change across the year

  Looking at figure 2, I can look to see if any health boards had significant changes in paracetamol prescriptions as the year progresses. I would perhaps have expected to observe greater paracetamol prescriptions in the winter months when compared to the rest of the year due to the greater number of illnesses such as viruses and colds occurring in these months [3]. However this does not appear to be the case when looking at figure 2 as I can't observe a significant increase on any of the graphs for the winter months. Most health boards appear to have at least a small peak in May when compared to the months either side as well as October. However this is only a slight increase and i am unable to think of any reason this would occur. The fact that most health boards are relatively consistent could suggest that most paracetamol prescriptions are for conditions that occur equally throughout the year. This could also point to a large amount of prescriptions being for chronic conditions and therefore equally spread throughout the year. Consequently, it would likely be beneficial in future to compare the health boards with the greatest prevalence of chronic conditions per population to which have highest prescriptions per person.

```{r Load Location Data, message = FALSE, results = 'hide'}
zip_path <- paste0("/vsizip//vsicurl/", "https://maps.gov.scot/ATOM/shapefiles/SG_NHS_HealthBoards_2019.zip")
hb_shapes <- st_read(zip_path) #reading geograohical data as saved online in a zip file
map_data <- hb_shapes %>% #adding geographical data to large dataframe
  left_join(full_prescriptions_info, by = c("HBCode" = "hbt"))

map_data_elderly <- map_data %>% #adding a column for proportion over 65
  mutate(over_65 = rowSums(across(86:111))) %>% 
  mutate(percent_over_65 = (over_65 / all_ages)*100) 
```
```{r interactive maps}
map_of_prescriptions <- map_data %>% #creating map of prescriptions per health board
   group_by(hb_name) %>% 
  summarise(number_of_paid_items = sum(number_of_paid_items), population = mean(all_ages), geometry = first(geometry)) %>% 
  mutate(items_per_100k = number_of_paid_items / population * 100000) %>% #calculating prescriptions per 100,000 people
  ggplot(aes(tooltip = hb_name, data_id = hb_name, fill = items_per_100k)) + #includes deciding what data shows when someone interacts with map
  geom_sf_interactive(colour = "black", size = 0.1) + #tells r to make map interactive
  scale_fill_distiller(palette = "Blues", direction = 1, name = "Prescriptions per 100k") +
  theme(plot.title = element_text(face = "bold", size = 10), plot.subtitle = element_text(size = 7))  +
  labs(title = "Figure 3a: Paracetamol Prescriptions in Scotland", subtitle = "Yearly Number of Paracetamol Prescriptions Per 100k in Each Health Board in Scotland (hover over areas to see health board name)") +
  theme(plot.title = element_text(face = "bold", size = 10), plot.subtitle = element_text(size = 7)) 

map_of_ages <- map_data_elderly %>% #creating map of proportion over 65 in each health board
   group_by(hb_name) %>% 
    summarise( number_of_paid_items = sum(number_of_paid_items), over_65 = mean(percent_over_65), geometry = first(geometry),) %>% #ensuring only one row per health board
  ggplot(aes(tooltip = hb_name, data_id = hb_name, fill = over_65)) + #includes deciding what data shows when someone interacts with map
  geom_sf_interactive( colour = "black", size = 0.1) + #tells r to make interactive map
  scale_fill_distiller(palette = "Blues", direction = 1, name = "Percentage over 65") +
  labs(title = "Figure 3b: Map of Older Populations in Scotland", subtitle = "Percentage of People 65 or Over in Each Health Board (hover over areas to see health board name)") +
  theme(plot.title = element_text(face = "bold", size = 10), plot.subtitle = element_text(size = 7))  

interactive_map_of_prescriptions <- girafe(ggobj = map_of_prescriptions) #converts the map to be interactive
interactive_map_of_ages <- girafe(ggobj = map_of_ages) #converts the map to be interactive
interactive_map_of_prescriptions #displays the prescriptions map
interactive_map_of_ages #displays the ages map
```
### Age and Paracetamol Prescriptions

  Using figures 3a and 3b, I can visualize if there is any correlation between paracetamol proportion of an areas population classified as elderly (over 65). The maps clearly show that Western Isles has both the greatest number of prescriptions per 100,000 people while also having the greatest proportion of its population over 65 years old. Additionally, the graph appears to show a reasonably strong correlation for Orkney, Dumfries & Galloway and Borders. All of which have high proportions of population over 65 while also having relatively high paracetamol prescriptions. On the other side, both Lothian and Greater Glasgow and Clyde have relatively low proportion of elderly people while also both having relatively low paracetamol prescriptions. Therefore, these health boards suggest that an older population may increase the amount of paracetamol prescriptions.
  However, one possible exception to this trend is Highland health board which appears to have a reasonably high elderly population while its paracetamol prescriptions are relatively low. Conversely, Shetland appears to have relatively high paracetamol prescriptions while not having a particularly large elderly population.
    Therefore, while it is clear from these maps that there is some correlation between age of a health board and it's paracetamol prescriptions, this isnt necessarily true for all boards. 
```{r SIMD}
simd <- url_loading("https://www.opendata.nhs.scot/dataset/78d41fa9-1a62-4f7b-9edb-3e8522a93378/resource/acade396-8430-4b34-895a-b3e757fa346e/download/simd2020v2_22062020.csv") %>% 
  rename(quintiles = simd2020v2country_quintile) %>% 
  select(quintiles, data_zone) #loading SIMD quintiles data per GP practise 

simd_gps <- url_loading("https://www.opendata.nhs.scot/dataset/f23655c3-6e23-4103-a511-a80d998adb90/resource/47557411-7eda-4278-9d6d-d26ed2ceab5a/download/practice_contact_details_20251001_o.csv") %>% 
  rename(gp_practice = practice_code) %>% 
  left_join(simd, by = "data_zone") %>% 
  filter(!(is.na(gp_practice)|is.na(quintiles))) #adds more information about the GP practices to SIMD data

simd_paracetamol <-full_prescriptions_info %>% 
  left_join(simd_gps, by = "gp_practice") #adds SIMD data and GP info to main data

simd_paracetamol_graph <- simd_paracetamol %>% #creating graph about SIMD and prescriptions
  group_by(gp_practice_name, date) %>% #to calculate prescriptions per patients separately
  summarise(items_per_100k = sum(number_of_paid_items / practice_list_size)*100000, quintiles = first(quintiles)) %>% 
  group_by(quintiles, date) %>% 
  summarise(mean_100k = mean(items_per_100k)) %>% 
  filter(!is.na(quintiles)) %>% 
  ggplot(aes(x = date, y = mean_100k)) +
  geom_line() +
  facet_wrap(~quintiles) +
   ggtitle("Figure 4: Average Paracetamol Prescriptions Per 100,000 people For GP Practices sorted by SIMD quintile") +
  xlab("Date") +
  ylab("Mean Prescriptions per 100,000") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1), plot.title = element_text(size=10), strip.text= element_text(size = 7)) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, NA), labels = scales::comma)
simd_paracetamol_graph #Displays SIMD graph
```

### Deprivation and Paracetamol Prescriptions

  Figure 4 shows how the SIMD quintile of GP practice affects average paracetamol prescriptions per 100,000 within these practices. For SIMD a lower number means a more deprived area (e.g 1 means most deprived). In the graph I can see that quintiles 2 and 3 have the highest mean paracetamol prescriptions per 100,000 patients, closely followed by the first quintile. There is then a visible decrease from the first three quintiles to the fourth quintile. The fifth quintile has a significantly lower mean prescriptions than the first four. This therefore fits my hypothesis to some extent from my introduction that the more deprived areas would be prescribed more paracetamol on average than the less deprived areas. This makes sense as people in deprived areas may have less money to spend on paracetamol themselves. Alternatively, they could be in deprived areas due to a chronic condition reducing their ability to work, with the condition possibly needing paracetamol as part of treatment, meaning the trend is due to chronic condition prevalence rather than SIMD quintile directly. However this relationship is weak as it is only really visible when going above the third quintile to the fourth and fifth. Therefore this finding has limited utility for guiding policy on paracetamol prescriptions as it is not a strong link and the link may be a correlation rather than causation.

## Limitations

  There are a few limitations reducing the validity of these results. Firstly I only used one years worth of data to create this analysis, meaning they do not account for variation between different years. Additionally, this analysis was very generalized, for the most part only looking at data at a health board level, meaning the results being averaged across the boards could have hidden more significant trends in smaller areas.

## Conclusions

  From this analysis I have observed that there are likely some demographics that receive more paracetamol prescriptions than others. Firstly, the islands appear to recieve more paracetamol prescriptions per 100,000 people while larger cities appear to receive less. I did not observe any significant differences in how paracetamol is prescribed across the different months. Age appeared to have a significant effect on prescriptions, with more prescriptions in places with a greater proportion of elderly residents. Finally, there appeared to be a trend in paracetamol prescriptions changing depending on SIMD quintile, however this was not a particularly significant trend. These findings may be useful in future to look at areas to target to reduce paracetamol prescriptions across Scotland, reducing financial burden on NHS. However, further research is required to confirm these findings and fully understand the causality of the observed trends.

## References

1.Find Out The Real Cost of 100 Paracetamol Tablets To The NHS! - Coastal Medical Partnership [Internet]. Coastalmedicalpartnership.nhs.uk. 2024. Available from: https://www.coastalmedicalpartnership.nhs.uk/find-out-the-real-cost-of-100-paracetamol-tablets-to-the-nhs

2.NHS England» Prescription curbs to free up hundreds of millions of pounds for frontline care [Internet]. www.england.nhs.uk. 2017. Available from: https://www.england.nhs.uk/2017/11/prescription-curbs-to-free-up-hundreds-of-millions-of-pounds-for-frontline-care/

3.LaMotte S. Scientists finally know why people get more colds and flu in winter [Internet]. CNN. 2022. Available from: https://edition.cnn.com/2022/12/06/health/why-winter-colds-flu-wellness